Design and implementation of a live coding environment for data science

### 2018-11-14 wip-live-programming/paper.tex
Rewriting the background section

\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{numclr}{rgb}{0.0,0.4,0.0}
\definecolor{strclr}{rgb}{0.4,0.4,0.0}
\definecolor{rstrclr}{rgb}{0.5,0.1,0.0}
\definecolor{prepclr}{rgb}{0.6,0.0,0.2}
\newcommand{\vect}[1]{\langl #1 \rangl}
\newcommand{\langl}{\begin{picture}(4.5,7)
\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\rangl}{\begin{picture}(4.5,7)
\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
\newcommand{\lsep}{~\,|\,~}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\rstr}[1]{\textnormal{\textcolor{rstrclr}{\sffamily "#1"}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\qident}[1]{\textnormal{\sffamily \guillemotleft #1\guillemotright}}
\newcommand{\dom}{\ident{dom}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}

@@

@@

\begin{figure*}
\includegraphics[scale=0.21]{wiki.png}
\includegraphics[scale=0.21]{drop.png}
\caption{Scraping 2016 science events. Preview of the Wikipedia source page (left) and live filtering (right).}
\label{fig:thegamma}
\end{figure*}

@@

\section{Live programming for data exploration}
\label{sec:live}

The Gamma aims to make basic data exploration accessible to non-experts, such as data journalists \cite{ddj}, 
who need to use data transparently. As such, we focus on simple scripts that can be written using
tools such as Jupyter notebooks \cite{jupyter}. Those scripts follow a typical data science workflow \cite{workflow};
they acquire and reformat data, run a number of analyses using the clean data and then create 
several visualizations.

An important property of data science workflow is that we work with readily available concrete data.
Data scientists load inputs into memory and refer to it in subsequent REPL interactions. They might 
later wrap completed working code into a function (and run on multiple datasets), but 
do not start with functions. We reflect this pattern in our language.

In this section, we provide a brief overview of The Gamma, a simple live coding environment for
data science. We discuss the language, type providers and the user interface,
before focusing on the algorithms behind live previews in Section~\ref{sec:formal}.

\subsection{The Gamma scripting language}
Figure~\ref{fig:thegamma} shows The Gamma script that scrapes items from a Wikipedia page, 
collects those marked as ``Events'' and filters them. The script illustrates two aspects of the
scripting language used by The Gamma -- its structure and its use of type providers for
dot-driven data exploration.

The scripting language is not intended to be as expressive as, say, R or Python and so it
has a very simple structure -- scripts are a sequence of let-bindings (that obtain or transform 
data) or statements (that produce visualizations). This reflects the fact that we always work
with concrete data and allows us to provide previews.

The most notable limitation of The Gamma is that the scripting language does not support top-level 
functions. This is not a problem for the simple scripts we consider, but it would be an issue for
a general-purpose language. We discuss potential design for functions that would still be based
on working with concrete data in Section~\ref{sec:design}.

\subsection{Dot-driven data exploration}

As illustrated by the second let binding in Figure~\ref{fig:thegamma}, many operations
in The Gamma can be expressed using member access via dot. The underlying mechanism is based
on type providers \cite{providers-fsharp,providers-idris}. The specific type provider used in 
the above example has been described elsewhere \cite{gamma}.

Given a data source (scraped Wikipedia page), the type provider generates type with members that allow
a range of transformations of the data such as grouping, sorting and filtering. Some of the members
are based only on the schema (e.g.~\ident{\textquotesingle Item Contains\textquotesingle} or 
\ident{\textquotesingle Drop Heading 2\textquotesingle}), but some may also be generated based
on (a sample of) the dataset (e.g.~the second member in \ident{\textquotesingle Heading 2 
is\textquotesingle.\textquotesingle Events[edit]\textquotesingle}).

What matters for the purpose of this paper is 
the fact that most operations are expressed via member access matters. First, it means 
that we need to provide live previews for sub-expressions formed by a chain of member accesses.
Second, it means that the result of any member access expression depends on the instance and,
possibly, on the parameters provided for a method call.

\subsection{Direct manipulation and live previews}
The screenshot in Figure~\ref{fig:thegamma} show the editor as implemented in The Gamma. This
includes live previews (discussed in this paper), but also an editor that provides spreadsheet-like
interface for editing the data transformation script.

In our implementation, previews appears below the currently selected let binding.
The preview can be of several types such as web page (left) or data table (right). In this paper,
we describe how we evaluate scripts to obtain the resulting object and we ignore how such objects
are rendered.

The Figure~\ref{fig:thegamma} also shows a special handling of expressions constructed using the
pivot type provider. The editor recognises individual data transformations and provides a simple
user interface for adding and removing transformations and changing their parameters that changes
the source code accordingly. This aspect is not discussed in the present paper.

>>>>

\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{numclr}{rgb}{0.0,0.4,0.0}
\definecolor{strclr}{rgb}{0.4,0.4,0.0}
\definecolor{rstrclr}{rgb}{0.5,0.1,0.0}
\definecolor{prepclr}{rgb}{0.6,0.0,0.2}
\newcommand{\vect}[1]{\langl #1 \rangl}
\newcommand{\langl}{\begin{picture}(4.5,7)
\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\rangl}{\begin{picture}(4.5,7)
\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
\newcommand{\lsep}{~\,|\,~}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\rstr}[1]{\textnormal{\textcolor{rstrclr}{\sffamily "#1"}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\qident}[1]{\textnormal{\sffamily \textquotesingle #1\textquotesingle}}
\newcommand{\dom}{\ident{dom}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}

@@

\section{Understanding how data scientists work}

Data scientists often use general-purpose programming languages such as Python, but the kind of 
code they write and the way they interact with the system is very different from how software 
engineers work.

This paper focuses on simple data wrangling and data exploration as done, for example, by 
journalists analysing government datasets. This section illustrates how such data analyses 
look and provides a justification for design decisions that we make when designing the data 
exploration calculus.

@@

\subsection{Data wrangling in Jupyter}
\label{sec:background-jupyter}

Increasing numbers of data analysts use notebook systems such as Jupyter or R Markdown, which 
make it possible to combine text, equations and code with results of running the code such as tables 
or visualizations. Notebooks blur the traditional distinction between development and execution. 
Data analysts write small snippets of code, run them immediately to see results and then revise 
them interactively based on the results. We aim to take this trend further and build foundations
for seamlessly integrating development and execution in a notebook-like environment.

Jupyter notebooks are used by a variety of users ranging from scientists who implement complex
models to journalists who load data, perform simple aggregations and create visualizations. Our
focus is on simple use cases, such as those of journalists, because we believe that this is where
better tooling can provide the most value.

As an example, consider the data anlysis done by the Financial Times for an article on plastic 
waste. The analysis joins datasets from Eurostat, UN Comtrade and more, aggregates the data and 
builds a visualization comparing the plastic waste flows in 2017 and 2018.
Figure~\ref{fig:ft-uncomtrade} shows an extract from one notebook from the data analysis. 
The code of the analysis has a number of remarkable properties:
%
\begin{itemize}
\item There is no abstraction. The analysis uses lambda functions as arguments to library calls,
  but it does not define any functions. Parameterization is done either by defining a list of 
  inputs and then having a for loop, or by defining a top-level variable and setting its value.
  In our example, \ident{material} is set to \str{plastics} and a comment lists other options. 
  This lets the analyst easily see and check the results of intermediate steps of the analysis.

\item The code is structured as a sequence of commands. Some commands define a variable, either by
  loading data, or by transforming data loaded previously. Even in Python, data is often treated
  as immutable. Other commands produce an output that is displayed in the notebook.

\item There are many corner cases, such as the fact that the \ident{keep\_default\_na} parameter 
  needs to be set to handle Namibia correctly. These are discovered interactively by writing and 
  running a code snippet and examining the output, so providing a rapid feedback is essential.
\end{itemize}

Many Jupyter notebooks are more complex than the example discussed here. They might define
helper functions or even structure code using object-oriented programming. However, simple 
data analyses such as the one discussed here are frequent enough to pose an interesting and
important problem for programming tools research.

\begin{figure}
\includegraphics[scale=0.6]{notebook.png}
\caption{An excerpt from a Jupyter notebook by Financial Times reporters, which loads trade
  data from the UN trade database and adds ISO country codes.}
\label{fig:ft-uncomtrade}
\end{figure}

\begin{figure}
\includegraphics[scale=0.52]{gamma.png}
\caption{The Gamma script to aggregate waste data. We select three countries and group
data by country. We then count the number of distinct reporting periods per country and type `.'
to choose further aggregation operations.}
\label{fig:gamma-waste}
\vspace{-1em}
\end{figure}

\subsection{Dot-driven data exploration in The Gamma}
\label{sec:background-gamma}

Data exploration of the kind discussed in the previous section has been the motivation
for a recent programming language research project The Gamma. The scripting language in The Gamma
captures the structure discussed above.

A script is a sequence of commands that can either define a 
variable or produce an output. The language does not support top-level function declarations, 
although we propose an extension that provides a simple abstraction mechanism and is compatible with
providing live previews in Section~X. Lambda functions can be used only as method 
arguments.

Given the expressiveness of The Gamma, libraries for it are implemented in other languages, such as
JavaScript. More importantly, The Gamma supports type providers, which can be used to generate types,
on the fly, for accessing and exploring external data sources. Type providers provide object types 
with members and The Gamma makes using those convenient by providing auto-complete when the user 
types the dot symbol (`.') to access a member.

The combination of type providers and auto-complete on~``.'' makes it possible to complete a 
large number of data exploration tasks through a very simple interaction of selecting operations from
a list. An example in Figure~\ref{fig:gamma-waste} summarizes waste data using a type provider.
The names of identifiers such as \qident{average Waste (kg)} do not have
any special meaning -- they are merely members generated by the type provider, based on information
about the data source. The type provider used in this example generates an object with members for
different data transformations, such as \qident{group data}, which return furhter objects with 
members for specifying transformation parameters, such as selecting the grouping key using
\qident{by Country}.

The Gamma is more complex than shown here -- it supports methods that can take
functions as arguments -- but the above example illustrates the fact that non-trivial data
exploration can be done using a very simple language.

We complement our theoretical work on foundations of live programming environments for data
exploration with an implementation for The Gamma. As discussed in the previous
section, the assumptions about structure of code that are explicit in The Gamma are also
implicitly present in Python and R data analyses produced by journalists, economists and
other users with other than  programming background. Consequently, our theoretical work 
applies to simple data analyses using Python and readers not familiar with The Gamma
can assume that we use a small subset of Python.

@@

\section{Calculus}

x

\begin{equation*}
\begin{array}{lcl}
p &=& c_1; \ldots; c_n\\
c &=& \kvd{let}~x = e ~|~ e\\
e &=& n ~|~ x ~|~ e.m(a, \ldots, a)\\
a &=& e ~|~ \lambda x\rightarrow e
\end{array}
\end{equation*}
